---
title: "PUBH 7462 - Homework 1"
author: "Maykon da Silva"
date: "1/30/2022"
output: github_document
html_document: default
pdf_document: default
---

```{r setup, include = FALSE}
#Load all the good stuff
library(tidyverse)
library(forcats)

#Working directory for .RMD
knitr::opts_knit$set(echo = TRUE,
                     root.dir = rprojroot::find_rstudio_root_file())

#Controlling figure output in markdown
knitr::opts_chunk$set(
#  fig.height =   
  fig.width = 6,
#  fig.asp = .5,
  out.width = "90%",
#  out.height = 
 fig.align = "center",
  cache = FALSE,
  echo  = TRUE
)

#Set Theme for ggplot2 - centers title and legend at bottom by default
theme_set(theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom"))

#Set Scientific notation output and decimal places for knitr
options(scipen = 999)
options(digits = 4)
```

## Problem 2.1 - Independent Bivariate Normal Random Sample

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Creating a tibble with 3 variables
tibble_hw1 <- tibble(
  x = rnorm(n = 1000, mean = 0, sd = 1),
  y = rnorm(n = 1000, mean = 1, sd = 2),
  sum_indicator = ifelse(test = x + y > 0.5, yes = TRUE, no = FALSE)
)

# Mutating "sum_indicator" variable to "Yes" and "No" and creating a ggplot
tibble_hw1 %>%
  mutate(sum_indicator = factor(ifelse(sum_indicator, "Yes", "No"))) %>%
  ggplot(aes(x = x, y = y, colour = fct_relevel(sum_indicator, "Yes", "No"))) +
  geom_point(alpha = 0.3,
             size  = 2,
             shape = 16) +
  geom_smooth(formula = y ~ x,
              method  = "loess",
              se      = FALSE) +
  scale_colour_viridis_d("X + Y > 0.5", option = "turbo") +
  labs(title = "Independent Bivariate Normal Random Sample") +
  xlab("X ~ N(0, 1)") +
  ylab("Y ~ N(1, 2)")
```

## Problem 2.2 - Penguin EDA

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Read data with relative path
penguin.df <- read_rds("./data/penguin.RDS")
```

### Problem 2.2.1 - Data Description

The "penguins" data set, found in the package "palmerpenguins" describes size measurements for adult foraging penguins near Palmer Station, Antarctica. More specifically, it includes measurements such as flipper length (mm), body mass (g), bill length (mm), and bill depth (mm) for three species (Adelie, Chinstrap, and Gentoo) and their respective sex (female and male) from three different islands (Biscoe, Dream and Torgersen). Observations for each of these variables were obtained from three years of study: 2007, 2008, and 2009. Therefore, the data set contains:

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Determining the number of rows and columns from the penguins data set
penguin_rows <- nrow(penguin.df)
penguin_columns <- ncol(penguin.df)
```

- `r penguin_columns` Variables: species, island, bill length (mm), bill depth (mm), flipper length (mm), body mass (g), sex, and year. 

- `r penguin_rows` observations/rows: each observation is a combination of all 8 variables and their levels (as described above). As an example, row 1 describes the bill length (39.1 mm), bill depth (18.7 mm), flipper length (181 mm) and body mass (3750 g) for a male Adelie penguin located in the Torgersen island in 2007. 

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Calculating the mean and standard deviation for flipper length and bill length
penguin.df_summary <- penguin.df %>%
  summarise(mean_flipper.length = mean(flipper_length_mm, na.rm = TRUE),
            mean_bill.length = mean(bill_length_mm, na.rm = TRUE),
            sd_flipper.length = sd(flipper_length_mm, na.rm = TRUE),
            sd_bill.length = sd(bill_length_mm, na.rm = TRUE)) %>%
  round(2)
```

The mean and standard deviation of flipper and bill length (mm) for these data set are as follows:

- Mean flipper length (mm) = `r penguin.df_summary$mean_flipper.length`

- Mean bill length (mm) = `r penguin.df_summary$mean_bill.length`

- Standard deviation of flipper length = `r penguin.df_summary$sd_flipper.length`

- Standard deviation of bill length = `r penguin.df_summary$sd_bill.length`


### Problem 2.2.2 Visualization

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Scatter plot of flipper length (y) by bill length (x), coloured by species
penguin.df %>%
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm, colour = species)) +
  geom_point(alpha = 0.6,
             size = 2,
             shape = 16) +
  geom_smooth(formula = y ~ x,
              method  = "lm",
              span    = 1,
              se      = FALSE) +
  labs(title = "Relationship between Flipper length and Bill length by Species") +
  ylab("Flipper Length (mm)") +
  xlab("Bill Length (mm)") +
  scale_colour_viridis_d("Species", option = "turbo")
```

```{r, echo = TRUE, warning = FALSE, error = FALSE, results = TRUE}

# Scatter plot of flipper length (y) by bill length (x), coloured by species and separated by sex
penguin.df %>%
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm, colour = species)) +
  facet_wrap(~sex) +
  geom_point(alpha = 0.6,
             size  = 2,
             shape = 16) +
  geom_smooth(formula = y ~ x,
              method  = "lm",
              span    = 1,
              se      = FALSE) +
  labs(title = "Relationship between Flipper length and Bill length by Species given Sex") +
  ylab("Flipper Length (mm)") +
  xlab("Bill Length (mm)") +
  scale_colour_viridis_d("Species", option = "turbo")
```

As shown by the plot above, there is a strong positive association between flipper length and bill length for all species of males penguins. This means that an increase in flipper length can be expected for an increase in bill length and vice-verse. However, for female penguins, this association does not seem to be as strong as for males. In addition, a small negative association between these two variables can be seen for the Adelie species, which means that a decrease in flipper length can be expected for an increase in bill length. It is important to point out that these relationships are not 1:1. A correlation analysis such as Pearson would present a better summary of the association between these two variables. 
